Implementation of Full Synchronous Composition Using IEC 61499 Function Blocks

### 2005-02-01 doc/Sync_Execution_with_IEC61499/paper.tex
*** empty log message ***

\begin{document}
\title{Implementation of Full Synchronous Composition of
  Discrete Event Models Using IEC 61499 Function Blocks}

@@

Shorter product life-cycles increase the demands on the
manufacturing system. Since the product, to be produced, in
the manufacturing plant is changing frequently the
production equipment needs to be modified too. The frequent
changes of the system make it time consuming and expensive
to update the control system application to reflect the
changes. The same problem applies to both system types
mentioned above and it becomes even larger in mission
critical control applications since the strict requirement
of guaranteeing, the correct operation is imposed. Thus,
there is a need to introduce algorithms and tools to make it
simple and as safe as possible to modify the control system.
The supervisory control theory,\cite{rw:con:1989}, is a
promising framework that might be useful for this purpose.

% supervisory control
Many man-made systems can, at some abstraction level, be
modeled by discrete event models. A useful class of discrete
event models is the deterministic finite automata which can
be used in supervisory control theory. In the theory these
models are used, together with a formal specification of the
desired behavior, to automatically generate a control
function which makes the system behave within specification.

% synchronization
The interaction between models is modeled by synchronization
operator. This operator has to be implemented in control
application in order to transfer theoretical results into
operating control system. The implementation has to be
distributable for it to be useful in control systems for the
applications distributed over geographical area, like the
ones mentioned in the beginning of the introduction.

There are existing methods for implementation of
synchronization operator using PLC programming techniques
\cite{hfl:syn:1999,hfl:mod:2001,fh:plc:1998,a:met:2002} but
the distribution is hard to achieve in that programming
environment. A new method presented in this paper uses the
emerging standard for distributed control systems to remedy
that.

% paper organization
Mathematical definitions that are used in the paper are
presented in the next section. After that, the most
important definitions and results of supervisory control
theory are introduced followed by an example showing how
supervisory control theory can be applied in implementation
of control application for a small system. After the example
a method for implementation of the synchronization operator
for any number of synchronized models is presented together
with the argument for the correctness of the implementation.
In Section~\ref{sec:app_SCT} some considerations for
implementation of supervisory control theory in applications
are presented before the conclusion of the paper.

\section{Mathematical Preliminaries}
In~\cite{hmu:int:2003} many important results for automata
are presented. Below, the terminology in this paper is
introduced.\\
\textbf{Definition:} Deterministic Finite Automaton (DFA)\\
A deterministic finite automaton is a 5-tuple defined as
$${A} = \langle Q, \Sigma, \delta, q_i, Q_m\rangle$$
where
$Q$ is a nonempty finite set of \emph{states}, $\Sigma$ is
the \emph{alphabet} which is a finite set of symbols that
denote the occurrence of an event, $\delta:Q\times \Sigma
\rightarrow Q$ is a \emph{transition function}, $q_i \in Q$
is the \emph{initial state} and $Q_m \subseteq Q$ is a set
of \emph{marked states}. $\Box$\\
\textbf{Definition:} Enabled Event Function\\
The enabled event function $\Gamma$ is defined as
$$\Gamma:Q\rightarrow 2^{\Sigma}$$
where $\Gamma(q)$ is the
set of all events $\sigma$ for which $\delta(q,\sigma)$ is
defined. $\Box$\\
Automata are usually visualized by directed graphs with
states represented by nodes, transitions by arrows and
events upon which the transitions occur by labels on the
arrows.

Synchronous composition, \cite{h:com:1985}, is a useful
approach to model the interaction between automata and is
used in this paper to model the interaction between the
system models.\\
\textbf{Definition}: Full Synchronous Composition of two DFAs\\
The full synchronous composition of two DFAs, $A^1$ and
$A^2$, is denoted ${A}^1||{A}^2=\langle
Q,\Sigma,\delta,q_i,Q_m\rangle$ where $Q=Q^1 \times Q^2$,
$\Sigma=\Sigma^1 \cup \Sigma^2$, $q_i=\langle {q^1_{i}},
{q^2_{i}} \rangle$, $Q_m={Q^1_{m}} \times {Q^2_{m}}$ and the
transition function is
\begin{displaymath}
  \begin{array}{l}
    \delta(\langle{q^1,q^2}\rangle,\sigma)=\\
    =\left\{ {\begin{array}{*{10}l}
          {{ \langle \delta ^1 (q^1 ,\sigma ), \delta ^2 (q^2 ,\sigma ) 
              \rangle}} &  \sigma \in \Gamma^1(q^1) \cap \Gamma^2(q^2) \\
          {{ \langle \delta ^1 (q^1 ,\sigma ), q^2 \rangle} } &  \sigma \in 
          \Gamma^1(q^1) \setminus \Sigma^2 \\
          { { \langle q^1, \delta ^2 (q^2 ,\sigma )\rangle}} & \sigma \in 
          \Gamma^2(q^2) \setminus \Sigma^1 \\
          { \rm undefined} & {{\rm otherwise.}}
        \end{array}} \right.
  \end{array}
\end{displaymath} 
where $q^1 \in Q^1$, $q^2 \in Q^2$ and $\sigma \in \Sigma$.
$\Box$ In other words, if an event is in both alphabets and
it is enabled in both automatons, the transitions associated
with it are made simultaneously in both automatons.
Otherwise, only the automaton containing the event in its
alphabet makes the transition. This definition can be
extended for several automata.\\
\textbf{Definition}: Full Synchronous Composition of $n$ DFAs\\
The full synchronous composition of DFAs, $A^1$, $A^2$ ...
$A^n$, is denoted ${A}^1||...||{A}^n=\langle
Q,\Sigma,\delta,q_i,Q_m\rangle$ where
\begin{eqnarray}
  Q & = & Q^1 \times Q^2 \times ... \times Q^n\nonumber\\
  \Sigma & = & \Sigma^1 \cup \Sigma^2 \cup ... \cup \Sigma^n\nonumber\\
  q_i & = & \langle q^1_i,q^2_i,...,q^n_i \rangle\nonumber\\
  Q_m & = & Q^1_m \times Q^2_m \times ... \times Q^n_m\nonumber
\end{eqnarray}
and the transition function is
\begin{multline}\label{eq:delta}
  \delta(\langle{q^1,...,q^n}\rangle,\sigma)=\\
  =
  \begin{cases}
    {\rm undefined} &{\rm if\ } \exists A^k:\sigma\in\Sigma^k\land\sigma\notin\Gamma^k(q^k)\\
    \langle q^{1+},\dots,q^{n+}\rangle & {\rm otherwise}
  \end{cases} 
\end{multline} 
where 
\begin{equation}\label{eq:qi}
  q^{k+}=
  \begin{cases}
    \delta^k(q^k,\sigma) & {\rm if\ }\sigma\in\Gamma^k(q^k)\\
    q^k                  & {\rm if\ }\sigma\notin\Sigma^k
  \end{cases}
\end{equation}
$\Box$\\
What this definition states is that in a group of automatons
that share one event if there is one automaton that does not
have that event enabled in its current state, than the event
is not allowed to occur in any of the other automatons in
the group. That is, one automaton is able to prevent all the
others from executing an event if it does not enable it at
the same time as all the other automata. This property of
the definition is used to construct a supervisor which
prevents events from occurring. The concept of supervisor is
explained further in the next section which introduces the
supervisory control theory.

@@

The supervisor, \texttt{S} (see Fig.~\ref{fig:sup_plant}),
has to prevent the plant, \texttt{P}, from breaking the
specification and from getting blocked in a state from where
it can not reach a marked state by using as few restrictions
on the plant as possible. By observing the observable events
of the plant and disabling the controllable events of the
plant, the supervisor decides when, and if, it has to limit
the activity of the plant in order to keep the plant's
behavior within the specification. An important result of
the supervisory control theory is that such a supervisor can
be calculated from the models of the plant and specification
so that
$$S||P$$
gives the system behavior.

In cases when there are several plant models $P_1,..,P_n$
with given specification models $Sp_1,...,Sp_n$ the
supervisor can also be generated and
$$S||P_1||...||P_n$$
gives the system behavior within specification.

Thus to control the system so that it satisfies the
specification, the synchronous composition of the plant and
supervisor models has to be implemented in the control
software. Using the conventional implementation methods,
i.e. implementing the synchronization using PLC programming
techniques, the solution becomes complex, which makes it
difficult to distribute across multiple control computers.
Next section presents an example illustrating how the
synchronous composition might be implemented using the new
standard.

@@

Consider a system containing two plants, \texttt{P1} and
\texttt{P2}, shown in Fig.~\ref{fig:plants}.These plant
models may represent two manufacturing resources with a
1-place buffer in-between. All events in the models are
controllable and observable. The desired behavior of the
system is that event \texttt{b} must always precede event
\texttt{c}, which means that a product must leave buffer
before another one can enter. The supervisor for this
specification can be generated and is shown in
Fig.~\ref{fig:sup}. In cases where the specification is
more complex it needs to be modeled using the automaton and
provided for the supervisor generation.
\begin{figure}[b]
  \centering 
  \fig[scale=0.55]{\texttt{P1}}{figures/P1_aut} 
  \fig[scale=0.55]{\texttt{P2}}{figures/P2_aut}
  \caption{Discrete event models of the plants. These may
    represent two manufacturing resources with a 1-place
    buffer in-between.}
  \label{fig:plants}
\end{figure}
\begin{figure}[b]
  \centering 
  \fig[scale=0.55]{\texttt{S}}{figures/Sp_aut}
  \caption{The model of the supervisor that make shure that
    event \texttt{b} always occurs before event \texttt{c}.}
  \label{fig:sup}
\end{figure}
\begin{figure}[t]
  \centering 
  \fig[scale=0.45]{\texttt{SYS}}{figures/Sys_aut}
  \caption{The model of possible system behavior while satisfying
    the desired system behavior, $S||P_1||P_2$.}
  \label{fig:sys}
\end{figure}
Full synchronous execution of these models gives the system
that behaves as freely as possible while fulfilling the
specification. The resulting model of the system is shown in
Fig.~\ref{fig:sys}.

% IEC 61499 intro
The emerging standard IEC 61499 specifies a generic
architecture for measurement and control of industrial
processes. The architecture is based on functional software
units called function blocks. The most common type of
function block in an application is basic function block and
it is shown in Fig.~\ref{fig:61499_intro} using the
graphical representation. This type of block executes
algorithms based on the arriving events and generates new
events that are passed on when the algorithms finish
execution. The algorithms use data associated with incoming
events to update the internal variables and produce the
output data associated with the outgoing events. Which
algorithms are invoked is controlled by the execution
control chart (ECC) which is an automaton with algorithms
associated with different states. When a state is entered
the algorithms associated with it are executed once and the
ECC stays in the state until a condition for leaving is
fulfilled. The conditions upon which transitions occur are
boolean expressions involving event input variables,
internal variables, input variables, and output variables.
When the expression evaluates to \texttt{true} the
transition occur. Special case is transition labeled with
\texttt{1}, it is taken as soon as the algorithms finish
execution.
\begin{figure}
  \centering
  \includegraphics[scale=0.65]{figures/basic_block}
  \caption{Anatomy of a basic function block type.}
  \label{fig:61499_intro}
\end{figure}

Function block types that generate the behavior of the plant
and supervisor models for the example are shown in
Fig.~\ref{fig:systemFBs}. The event input of each block type
is \texttt{OCCURRED}. It signals that an discrete event has
occurred and that the model should update its state
according to the data inputs. The data inputs are boolean
variables representing the discrete events in the model's
alphabet. For well defined behavior only one of the data
inputs associated with the \texttt{OCCURRED} event must be
\texttt{true} when the event is received. The model's state
is then updated according to the discrete event that this
data input represents and the event output \texttt{DONE} is
generated. It signals that the model has updated its state
and that the data outputs are ready. The data outputs are
boolean variables representing the enabled events after the
model has updated its state, i.e. the events that are
possible from the new state.

The function block type that is generated according to the
new method presented is shown in Fig.~\ref{fig:syncFB} and
is called synchronization function block type. The data
inputs and outputs of the block type are application
specific, since they are defined by the models that will be
synchronized. The event input and output, on the other hand,
are the same for all applications which also makes the ECC
of the block type the same for all applications. The ECC for
the block type is shown in Fig.~\ref{fig:ecc_sync}. The
\texttt{DISABLE\_EVENTS} algorithm associated with the
\texttt{SYNC} state is where the synchronized execution is
implemented. The \texttt{UPDATE} event is sent as soon as
the \texttt{DISABLE\_EVENTS} algorithm has finished
execution.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{figures/P1}\\(a)
  \vskip 5mm
  \includegraphics[scale=0.45]{figures/P2}\\(b)
  \vskip 5mm
  \includegraphics[scale=0.45]{figures/Sp}\\(c)
  \caption{The function block types implementing (a)
    plant 1, (b) plant 2, and (c) supervisor.}
  \label{fig:systemFBs}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[scale=0.5]{figures/SYNC}
  \caption{The function block type for synchronization.}
  \label{fig:syncFB}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[scale=0.45]{figures/sync_ecc}
  \caption{The execution control chart for the
    synchronization function block type.}
  \label{fig:ecc_sync}
\end{figure}

The synchronization block reads the enabled events of the
synchronized models and disables the events that are not
enabled in all the models that have that event in their
alphabets. Each boolean data input signals if the event in
the model represented by that input is enabled
(\texttt{true}) or disabled (\texttt{false}). By setting the
corresponding data output to \texttt{false} the
synchronization block signals that the transition labeled by
that event is not allowed to occur in the model during the
next update cycle. To calculate each data output a logical
expression consisting of all data inputs that represent the
same event, connected by logical \texttt{and} operation, is
used. This means that if the event is enabled in all of the
models that have the event in their alphabet it is allowed
to occur in the next update cycle. Otherwise the expression
is false and the event is disabled. Following implementation
of the \texttt{DISABLE\_EVENTS} algorithm in IEC 61131
structured text does the computation:
\begin{verbatim}
EO_P1_a := EI_P1_a;
EO_P1_b := EI_P1_b AND EI_S_b;
EO_P2_c := EI_P2_c AND EI_S_c;
EO_P2_d := EI_P2_d;
EO_S_b := EI_P1_b AND EI_S_b;
EO_S_c := EI_P2_c AND EI_S_c;
\end{verbatim}
Since the event \texttt{a} is in the alphabet of the
\texttt{P1} model only, it is always allowed to occur if it
is enabled. The event \texttt{b} is allowed to occur in
\texttt{P1} model only if it is enabled in \texttt{P1} and
\texttt{S} models at the same time since it is contained in
both alphabets. The same applies to the event \texttt{b} in
\texttt{S} model, hence the same logical expression. Events
\texttt{d} and \texttt{c} are treated in the same manner.

@@

When algorithm is generated what is left is to connect
instances of the model function blocks with the
synchronization function block to generate the control
application. Some further considerations for the
implementation of supervisory control theory are presented in
the following section.

\section{Applications in supervisory control theory}
\label{sec:app_SCT}
Since the general method presented is suitable for computer
implementation it is interesting to investigate how this
fact can ease the implementation of control applications
using supervisory control theory.

@@

In software tool Supremica \cite{aff:sup:2003,www:sup:2004},
algorithms from supervisory control theory are implemented
so that a supervisor can be synthesized from automata models
of the plants and specifications. The algorithm that
implements the general method for synchronization according
to this paper is also implemented in the same tool. What
this means is that the control engineer can focus on the
correct modeling of the plants and transferring of the
informal specifications to the specification models for use
in the Supremica tool. Then the tool automatically generates
the supervisor in form of an automaton model, function block
types that implement the skeleton for the plants, the
supervisor automaton, synchronous composition function block
and finally the control application that connects the
instances of the generated function block types as seen in
Fig.~\ref{fig:app} for the example. The skeleton in this
case refers to the fact that the communication with the
hardware has to be implemented by hand but all the behavior
of the models is automatically generated. Using the tool to
generate the control application gives new application
structure as shown in Fig.~\ref{fig:new_app_struct}. In
this structure, only the models for the generation of the
plant blocks and the hardware control portion of those
blocks have to be upgraded whenever the plants in the system
change. All other parts of the application are regenerated
automatically.
\begin{figure}[b]
  \centering 
  \includegraphics[scale=0.6]{figures/new_app_struct}
  \caption{A new structure for distributed control applications.}
  \label{fig:new_app_struct}
\end{figure}

When trying to implement the control application for complex
systems with many plants and specifications the problem of
state space explosion is encountered when trying to
synthesize the supervisor model. There are methods that
represent the plants, specifications and the supervisor in
different ways to try to cope with the problem. For example,
a supervisor may be expressed as a single automaton,
multiple automata, and a binary decision diagram for
off-line synthesis. Or the supervisor may be synthesized
on-line while the control application is executing. The
method presented in this paper can be used for
synchronization of all types of representations since the
model interface (see Fig.~\ref{fig:model_interface}) is
based on events which are a general characteristic of the
discrete event models.

\section{Conclusion}
Using supervisory control theory in construction of control
applications can significantly reduce development time but
until now it was difficult to apply the theory to
implementation of distributed control applications. Using
the emerging standard IEC 61499 it becomes significantly
easier. The method presented in this paper brings
supervisory control theory together with distributed control
application development in an efficient way.

Since the method is implemented in the Supremica tool it
significantly contributes to the shorter control application
development time since it is a part in automatic generation
of control applications. The method is also future proof by
using an inherent characteristic of discrete event models
for the interface upon which synchronization function block
is generated.

The method and function block programming in general can be
tested using the Function Block Development Kit (FBDK),
\cite{c:fun:2002}. For further information about IEC 61499
function blocks see \cite{c:ope:2002}.

>>>>

\begin{document}
\title{Implementation of Full Synchronous Composition Using
  IEC 61499 Function Blocks}

@@

Shorter product life-cycles increase the demands on the
manufacturing system. Since products to be produced in
manufacturing plants are changing frequently, the production
equipment also needs to be modified. Frequent changes of the
system make it time consuming and expensive to update the
control system application to reflect the such changes. The
problem becomes even larger in mission critical control
applications since strict requirement of guaranteeing
correct operation is then imposed. Thus, there is a need to
introduce algorithms and tools to make it simple and as safe
as possible to modify the control system. The supervisory
control theory,\cite{rw:con:1989}, is a promising framework
that might be useful for this purpose.

% supervisory control
Many man-made systems can, at some abstraction level, be
modeled by discrete event models. A useful class of discrete
event models is the deterministic finite automata, which can
be used in supervisory control theory. In the theory these
models are used together with a formal specification of the
desired behavior to automatically generate a control
function which makes the system behave within the given
specification.

% synchronization
The interaction between subsystems is modeled by Hoare's
full synchronous composition (FSC)~\cite{h:com:1985}. This
operator has to be implemented in control applications in
order to transfer theoretical results into operating control
systems. The implementation has to be distributable to be
applicable in control systems distributed over geographical
areas, like the ones mentioned above.

There are existing methods for implementation of FSC using
PLC programming techniques
\cite{hfl:syn:1999,hfl:mod:2001,fh:plc:1998,a:met:2002} but
the distribution is hard to achieve in that programming
environment. A new method presented in this paper uses the
emerging standard for distributed control systems to remedy
that.

% paper organization
Mathematical definitions that are used in the paper are
presented in the next section. Important definitions and
results of supervisory control theory are introduced,
followed by an example showing how supervisory control
theory can be applied in implementation of control
application for a small system. After the example a method
for implementation of FSC for any number of synchronized
models is presented together with the argument for the
correctness of the implementation. Finally some
considerations for implementation of supervisory control
theory in applications are presented before the conclusion
of the paper.

\section{Mathematical Preliminaries}
In~\cite{hmu:int:2003} many important results for automata
are presented. Below, the terminology in this paper is
introduced.

\textbf{Definition 1:} Deterministic Finite Automaton (DFA)\\
A deterministic finite automaton is a 5-tuple defined as
$${A} = \langle Q, \Sigma, \delta, q_i, Q_m\rangle$$
where
$Q$ is a nonempty finite set of \emph{states}, $\Sigma$ is
the \emph{alphabet} which is a finite set of symbols that
denote the occurrence of an event, $\delta:Q\times \Sigma
\rightarrow Q$ is a \emph{transition function}, $q_i \in Q$
is the \emph{initial state} and $Q_m \subseteq Q$ is a set
of \emph{marked states}.\hfill$\Box$

\textbf{Definition 2:} Enabled Event Function\\
The enabled event function $\Gamma$ is defined as
$$\Gamma:Q\rightarrow 2^{\Sigma}$$
where $\Gamma(q)$ is the
set of all events $\sigma$ for which $\delta(q,\sigma)$ is
defined.\hfill$\Box$\\
Automata are usually visualized by directed graphs with
states represented by nodes, transitions by arrows and
events upon which the transitions occur by labels on the
arrows.

Synchronous composition, \cite{h:com:1985}, is a useful
approach to model the interaction between automata and is
used in this paper to model the interaction between the
system models.

\textbf{Definition 3}: Full Synchronous Composition of two DFAs\\
The full synchronous composition of two DFAs, $A^1$ and
$A^2$, is denoted ${A}^1||{A}^2=\langle
Q,\Sigma,\delta,q_i,Q_m\rangle$ where $Q=Q^1 \times Q^2$,
$\Sigma=\Sigma^1 \cup \Sigma^2$, $q_i=\langle {q^1_{i}},
{q^2_{i}} \rangle$, $Q_m={Q^1_{m}} \times {Q^2_{m}}$ and the
transition function is
\begin{displaymath}
  \begin{array}{l}
    \delta(\langle{q^1,q^2}\rangle,\sigma)=\\
    =\left\{ {\begin{array}{*{10}l}
          {{ \langle \delta ^1 (q^1 ,\sigma ), \delta ^2 (q^2 ,\sigma ) 
              \rangle}} &  \sigma \in \Gamma^1(q^1) \cap \Gamma^2(q^2) \\
          {{ \langle \delta ^1 (q^1 ,\sigma ), q^2 \rangle} } &  \sigma \in 
          \Gamma^1(q^1) \setminus \Sigma^2 \\
          { { \langle q^1, \delta ^2 (q^2 ,\sigma )\rangle}} & \sigma \in 
          \Gamma^2(q^2) \setminus \Sigma^1 \\
          { \rm undefined} & {{\rm otherwise.}}
        \end{array}} \right.
  \end{array}
\end{displaymath} 
where $q^1 \in Q^1$, $q^2 \in Q^2$ and $\sigma \in \Sigma$.
\hfill$\Box$\\
In other words, if an event is in both alphabets and
it is enabled in both automata, the transitions associated
with it are made simultaneously in both automata.
Otherwise, only the automaton containing the event in its
alphabet makes the transition. This definition can be
extended for several automata.

\textbf{Definition 4}: Full Synchronous Composition of $n$ DFAs\\
The full synchronous composition of DFAs, $A^1$, $A^2$ ...
$A^n$, is denoted ${A}^1||...||{A}^n=\langle
Q,\Sigma,\delta,q_i,Q_m\rangle$ where
\begin{eqnarray}
  Q & = & Q^1 \times Q^2 \times ... \times Q^n\nonumber\\
  \Sigma & = & \Sigma^1 \cup \Sigma^2 \cup ... \cup \Sigma^n\nonumber\\
  q_i & = & \langle q^1_i,q^2_i,...,q^n_i \rangle\nonumber\\
  Q_m & = & Q^1_m \times Q^2_m \times ... \times Q^n_m\nonumber
\end{eqnarray}
and the transition function is
\begin{multline}\label{eq:delta}
  \delta(\langle{q^1,...,q^n}\rangle,\sigma)=\\
  =
  \begin{cases}
    {\rm undefined} &{\rm if\ } \exists A^k:\sigma\in\Sigma^k\land\sigma\notin\Gamma^k(q^k)\\
    \langle q^{1+},\dots,q^{n+}\rangle & {\rm otherwise}
  \end{cases} 
\end{multline} 
where 
\begin{equation}\label{eq:qi}
  q^{k+}=
  \begin{cases}
    \delta^k(q^k,\sigma) & {\rm if\ }\sigma\in\Gamma^k(q^k)\\
    q^k                  & {\rm if\ }\sigma\notin\Sigma^k
  \end{cases}
\end{equation}
\hfill$\Box$\\
What this definition states is that, in a group of automata
that share one event, if there is one automaton that does not
have that event enabled in its current state, than the event
is not allowed to occur in any of the other automata in
the group. That is, one automaton is able to prevent all the
others from executing an event if it does not enable it at
the same time as all the other automata. This property of
the definition is used to construct a supervisor which
prevents events from occurring. The concept of supervisor is
explained further in the next section which introduces the
supervisory control theory.

@@

The supervisor, \texttt{S} (see Fig.~\ref{fig:sup_plant}),
has to prevent the plant, \texttt{P}, from breaking the
specification and from getting blocked in a state, from where
it can not reach a marked state, by using as few restrictions
on the plant as possible. By observing the observable events
of the plant and disabling controllable events of the
plant, the supervisor decides when and if it has to limit
the activity of the plant in order to keep the plant's
behavior within the specification. An important result of
the supervisory control theory is that such a supervisor can
be calculated from the models of the plant and specification
so that
$$S||P$$
gives the desired system behavior with as few
restriction of the plant as possible.

In cases when there are several plant models $P_1,..,P_n$
with given specification models $Sp_1,...,Sp_n$ the
supervisor can also be generated and
$$S||P_1||...||P_n$$
gives the system behavior within the specification.

Hence, to control the system so that it satisfies the
specification, the FSC of the plants and supervisor models
has to be implemented in the control software. Using the
conventional implementation methods, i.e. implementing the
synchronization using PLC programming techniques makes it
difficult to distribute the application across multiple
control computers, which is needed when the plants are
distributed over geographical areas. Next section presents
an example illustrating how the synchronous composition
might be implemented using the new standard.

@@

Consider a system containing two plants, \texttt{P1} and
\texttt{P2}, shown in Fig.~\ref{fig:plants}. All events in
the models are controllable and observable. The desired
behavior of the system is that event \texttt{b} must always
precede event \texttt{c}. These plant models together with
the specification may represent two manufacturing resources
with a 1-place buffer in-between. The specification means
then that a product must leave buffer before another one can
enter. The supervisor for this specification can be
generated and is shown in Fig.~\ref{fig:sup}.
\begin{figure}[b]
  \centering 
  \fig[scale=0.55]{\texttt{P1}}{figures/P1_aut} 
  \fig[scale=0.55]{\texttt{P2}}{figures/P2_aut}
  \caption{Discrete event models of the plants. These may
    represent two manufacturing resources with a 1-place
    buffer in-between.}
  \label{fig:plants}
\end{figure}
\begin{figure}[b]
  \centering 
  \fig[scale=0.55]{\texttt{S}}{figures/Sp_aut}
  \caption{The model of the supervisor that makes sure 
    event \texttt{b} always occurs before event \texttt{c}.}
  \label{fig:sup}
\end{figure}
\begin{figure}[t]
  \centering \fig[scale=0.4]{\texttt{SYS}}{figures/Sys_aut}
  \caption{The model of possible system behavior while satisfying
    the desired system behavior, $S||P_1||P_2$.}
  \label{fig:sys}
\end{figure}
Full synchronous execution of \texttt{P1}, \texttt{P2} and
\texttt{S} gives the system that behaves as freely as
possible while fulfilling the specification. The resulting
model of the system is shown in Fig.~\ref{fig:sys}.

\subsection{IEC 61499 Basics}
The emerging standard IEC 61499 specifies a generic
architecture for measurement and control of industrial
processes. The architecture is based on functional software
units called function blocks. The most common type of
function block in an application is the basic function block
shown in Fig.~\ref{fig:61499_intro} using the graphical
representation. This type of block executes algorithms based
on the arriving events and generates new events that are
passed on when the algorithms finish execution. The
algorithms use data associated with incoming events to
update the internal variables and produce the output data
associated with the outgoing events. Which algorithms are
invoked is controlled by the execution control chart (ECC),
which is an automaton with algorithms associated with
different states. When a state is entered the algorithms
associated with it are executed once and the ECC stays in
the state until a condition for leaving is fulfilled. The
conditions upon which transitions occur are boolean
expressions involving event input variables, internal
variables, input variables, and output variables. When the
expression evaluates to \texttt{true} the transition occur.
Special case is transition labeled with \texttt{1}; it is
taken as soon as the algorithms finish execution.
\begin{figure}
  \centering
  \includegraphics[scale=0.8]{figures/basic_block}
  \caption{Anatomy of a basic function block type.}
  \label{fig:61499_intro}
\end{figure}

\subsection{Example Implementation}
Function block types that generate the behavior of the plant
and supervisor models for the example are shown in
Fig.~\ref{fig:systemFBs}. The event input of each block type
is \texttt{OCCURRED}. It signals that a discrete event has
occurred and that the model should update its state
according to the data inputs. The data inputs are boolean
variables representing the discrete events in the model's
alphabet. For well defined behavior only one of the data
inputs associated with the \texttt{OCCURRED} event must be
\texttt{true} when the event is received. The model's state
is then updated according to the discrete event that this
data input represents and the event output \texttt{DONE} is
generated. It signals that the model has updated its state
and that the data outputs are ready. The data outputs are
boolean variables representing the enabled events after the
model has updated its state, i.e. the events that are
possible from the new state.

The function block type that is generated according to the
new method presented is shown in Fig.~\ref{fig:syncFB} and
is called synchronization function block type. The data
inputs and outputs of the block type are application
specific, since they are defined by the models that will be
synchronized. The event input and output, on the other hand,
are the same for all applications which also makes the ECC
of the block type the same for all applications.

The synchronization block reads the enabled events of the
synchronized models and disables the events that are not
enabled in all the models that have that event in their
alphabets. Each boolean data input signals if the event in
the model represented by that input is enabled
(\texttt{true}) or disabled (\texttt{false}). By setting the
corresponding data output to \texttt{false} the
synchronization block signals that the transition labeled by
that event is not allowed to occur in the model during the
next update cycle.

The ECC for the block type is shown in
Fig.~\ref{fig:ecc_sync}. The \texttt{DISABLE\_EVENTS}
algorithm associated with the \texttt{SYNC} state is where
the synchronized execution is implemented. The
\texttt{UPDATE} event is sent as soon as the
\texttt{DISABLE\_EVENTS} algorithm has finished execution.
The following implementation of the algorithm in IEC 61131
structured text does the computation:
\begin{verbatim}
EO_P1_a := EI_P1_a;
EO_P1_b := EI_P1_b AND EI_S_b;
EO_P2_c := EI_P2_c AND EI_S_c;
EO_P2_d := EI_P2_d;
EO_S_b := EI_P1_b AND EI_S_b;
EO_S_c := EI_P2_c AND EI_S_c;
\end{verbatim}
To calculate each data output a logical expression
consisting of all data inputs that represent the same event,
connected by logical \texttt{and} operation, is used. This
means that if the event is enabled in all of the models that
have the event in their alphabet it is allowed to occur in
the next update cycle. Otherwise the expression is false and
the event is disabled.

Since the event \texttt{a} is in the alphabet of the
\texttt{P1} model only, it is always allowed to occur if it
is enabled. The event \texttt{b} is allowed to occur in
\texttt{P1} model only if it is enabled in \texttt{P1} and
\texttt{S} models at the same time since it is contained in
both alphabets. The same applies to the event \texttt{b} in
\texttt{S} model, hence the same logical expression. Events
\texttt{d} and \texttt{c} are treated in the same manner.

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{figures/P1}\\(a)
  \vskip 5mm
  \includegraphics[scale=0.5]{figures/P2}\\(b)
  \vskip 5mm
  \includegraphics[scale=0.5]{figures/Sp}\\(c)
  \caption{The function block types implementing (a)
    plant 1, (b) plant 2, and (c) supervisor.}
  \label{fig:systemFBs}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[scale=0.5]{figures/SYNC}
  \caption{The function block type for synchronization.}
  \label{fig:syncFB}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[scale=0.5]{figures/sync_ecc}
  \caption{The execution control chart for the
    synchronization function block type.}
  \label{fig:ecc_sync}
\end{figure}

@@

When the algorithm is generated what is left is to connect
instances of the model function blocks with the
synchronization function block to generate the control
application. Some further considerations for the
implementation of supervisory control theory are presented
in the following section.

\section{Applications in supervisory control theory}
\label{sec:app_sct}
Since the general method presented is suitable for computer
implementation it is interesting to investigate how this
fact can ease the implementation of control applications
using supervisory control theory.

@@

In the software tool Supremica
\cite{aff:sup:2003,www:sup:2004}, algorithms from
supervisory control theory are implemented so that a
supervisor can be synthesized from automata models of the
plants and specifications. The algorithm that implements the
general method for synchronization according to this paper
is also implemented in the same tool. What this means is
that the control engineer can focus on the correct modeling
of the plants and transferring of the informal
specifications to the specification models for use in the
Supremica tool. Then the tool automatically generates the
supervisor in form of an automaton model, function block
types that implement the skeleton for the plants, the
supervisor automaton, synchronous composition function block
and finally the control application that connects the
instances of the generated function block types as seen in
Fig.~\ref{fig:app} for the example. The skeleton in this
case refers to the fact that the communication with the
hardware has to be implemented by hand but all the behavior
of the models is automatically generated. Using the tool to
generate the control application gives new application
structure as shown in Fig.~\ref{fig:new_app_struct}. In this
structure, only the models for the generation of the plant
blocks and the hardware control portion of those blocks have
to be upgraded whenever the plants in the system change. All
other parts of the application are regenerated
automatically.

When trying to implement the control application for complex
systems with many plants and specifications the problem of
state space explosion is encountered when trying to
synthesize the supervisor model. There are methods that
represent the plants, specifications and the supervisor in
different ways to try to cope with the problem. For example,
a supervisor may be expressed as a single automaton,
multiple automata, and a binary decision diagram for
off-line synthesis. Or the supervisor may be synthesized
on-line while the control application is executing. The
method presented in this paper can be used for
synchronization of all types of representations since the
model interface (see Fig.~\ref{fig:model_interface}) is
based on events which are a general characteristic of the
discrete event models.
\begin{figure}[!hb]
  \centering 
  \includegraphics[scale=0.6]{figures/new_app_struct}
  \caption{A new structure for distributed control applications.}
  \label{fig:new_app_struct}
\end{figure}
\vfil

\section{Conclusion}
Using supervisory control theory in construction of control
applications can significantly reduce development time, but
until now it was difficult to apply the theory to
implementation of distributed control applications. Using
the emerging standard IEC 61499 it becomes significantly
easier. The method presented in this paper brings
supervisory control theory together with distributed control
application development in an efficient way.

Since the method is implemented in the Supremica tool it
significantly contributes to the shorter control application
development time, since it is a part in automatic generation
of control applications. The method is also future proof by
using an inherent characteristic of discrete event models
for the interface upon which synchronization function block
is generated.

The method and function block programming in general can be
tested using the Function Block Development Kit (FBDK),
\cite{c:fun:2002}. For further information about IEC 61499
function blocks, see \cite{c:ope:2002}.
